Early warning signals: The charted and uncharted territories

### 2013-05-21 manuscript.tex
Modified discussion of Hopf to reflect Alan's comments

Many natural systems exhibit regime shifts - rapid changes in the state
and conditions of system behavior. Such shifts include lake
eutrophication (Carpenter et al. 1999), algal overgrowth of coral
systems (Mumby et al. 2007), fishery collapse (Jackson et al. 2001),
desertification of grasslands (Kéfi et al. 2007), and rapid changes in
climate (Dakos et al. 2008, Lenton et al. 2009). Such dramatic shifts
have the potential to impact ecosystem health and human well-being.
Thus, it is important to develop strategies for adaptation, mitigation,
and avoidance of such shifts.

The idea that complex systems such as ecosystems could change suddenly
and without warning goes back to the 1960s (Lewontin 1969, Holling 1973,
May 1977). Such early work revealed that even simple models with the
appropriate nonlinearities were capable of unpredictable behavior. The
only way to predict the transition was to have the right model -- and
that meant having already had the chance to observe the transition (even
then, this remains a tough problem). One cogent early example (Ludwig et
al. 1978) demonstrated how knowledge of the forms and time scales of
interactions among insects, birds, and trees could lead to a qualitative
model that essentially predicted the possibility of regime shifts.

@@

Carpenter et al. (2011) provide a larger-scale example in which a lake
ecosystem is manipulated towards a sudden transition through the
introduction of a predator, while a neighboring experimental lake
provides a control. In this and similar lake systems, bifurcation is
thought to be driven in part by trophic interactions where adult fish
prey on the competitors of their juveniles (Carpenter and Kitchell 1996,
Walters2001; Carpenter et al. 2008) which leads to a saddle-node
bifurcation. While the underlying dynamics of a whole lake ecosystem are
less tractable than the laboratory controlled chemoststats of
microorganisms, the system is understood well enough to anticipate that
a sudden transition can be induced under the intended manipulation. Like
the laboratory examples, this helps eliminate the options outside the
circle ``bifurcations,'' in Figure 1. The observed warning signals then
place it in the center of the diagram.

@@

Not all regime shifts are rapid. Some systems undergo bifurcations
between qualitatively different, but quantitatively similar regimes.
These transitions may be reversible. In a management setting, such
qualitative changes may insignificant, so warning signals that detect
such transitions may be effective ``false positives.''

@@

Large, rapid changes in external conditions will result in rapid changes
in ecological system dynamics. For instance, rapid changes in North
American vegetation at the start of the Bølling-Allerød and end of the
Younger Dryas period are thought to be responses to similarly large,
rapid changes in climate (Williams et al. 2011). Doney and Sailley
(2013) interpret a recent analysis by Di Lorenzo and Ohman (2013) as
demonstrating that what were previously thought of as regime shifts in
krill dynamics in the Pacific ocean (Hare and Mantua 2000) could
actually be explained by a close coupling to the external forcing of El
Nino environmental dynamics through the Pacific Decadal Oscillation
(PDO). Schooler et al. (2011) found that lakes with the invasive plant
\emph{Salvaniai molesta} andherbivorous weevils alternated between low-
and high-\emph{Salvnia} states driven by disturbances from regular
external flooding events. These examples highlight cases that involve
critical transitions between regimes under circumstances that do not
permit the discovery of early warning signals, as CSD is not anticipated
under these mechanisms.

@@

A wide variety of statistical summary indicators have been examined as
potential detectors of CSD. The most common are variance and
autocorrelation. Others include skewness {[}Guttal2008a{]} and
conditional heteroscedasticity (Seekell et al. 2011). These statistics
are typically calculated on sliding windows of time-series data and
tested formally or informally for trends. The relative power of these
tests varies considerably with context; no indicator has consistently
outperformed others (Dakos et al. 2011b, 2012, Lindegren et al. 2012,
Perretti and Munch 2012). Also, measuring these indicators requires
making sometimes arbitrary calculations. For instance, the power of
lag-1 autocorrelation to detect a regime shift may be modified by
changing methods of data aggregation, de-trending, changing sliding
window length, filtering signal bandwidth (Lenton et al. 2012). These
choices may be optimized when enough calibration data is available, as
Lenton et al. (2012) were able to do with several sets of paleoclimate
data. However, such calibration may not be possible with many ecological
datasets. Multiple-method (Lindegren et al. 2012) and composite indices
(Drake and Griffen 2010) have been proposed, but their power relative to
other indicators is unknown.

@@

Thus, establishing the saddle node mechanism is a necessary condition of
using CSD as a warning signal. This can be done via manipulation in
simple experimental systems (Veraart et al. 2012, Dai et al. 2012), but
this is impractical in most natural systems. Another approach is to
assume the saddle-node mechanism applies to a limited set of systems
that have well-studied examples, such as lakes undergoing eutrophication
(Scheffer et al. 2001), lakes with `trophic-triangle' cascade mechanisms
(Carpenter and Kitchell 1996, Walters2001; Carpenter et al. 2008),
forest/savannah transitions (Staver et al. 2011, Hirota et al. 2011),
and rangeland transitions (Walker1993; Anderies et al. 2002). Fitting
simplified saddle-node models to past regime shifts (Boettiger and
Hastings 2012a) in less well-understood systems may provide evidence for
the mechanism. However, care must be taken to specify sufficient
alternative models.

@@

The future of early warning signals lies in the uncharted territory. For
certain classes of transitions, such as stochastically-driven regime
shifts, prediction may not be possible. In such cases, managing for
resilience may be the only option. Likewise, regime shifts driven by
external perturbation or strong forcing are not predictable \emph{if}
the scope of management does not include the external causes. Proper
scoping of the management problem can avoid this situation (Fischer et
al. 2009, Alliance 2010, Polasky et al. 2011). More research is needed
in methods of distinguishing such cases from those in which early
detection may be possible.

@@

Anderies, J. M., M. a Janssen, and B. H. Walker. 2002. Grazing
Management, Resilience, and the Dynamics of a Fire-driven Rangeland
System. Ecosystems 5:23--44.

@@

Barnosky, A. D., E. a Hadly, J. Bascompte, E. L. Berlow, J. H. Brown, M.
Fortelius, W. M. Getz, J. Harte, A. Hastings, P. a Marquet, N. D.
Martinez, A. Mooers, P. Roopnarine, G. Vermeij, J. W. Williams, R.
Gillespie, J. Kitzes, C. Marshall, N. Matzke, D. P. Mindell, E. Revilla,
and A. B. Smith. 2012. Approaching a state shift in Earth's biosphere.
Nature 486:52--58.

@@

Brock, W. a, and S. R. Carpenter. 2012. Early Warnings of Regime Shift
When the Ecosystem Structure Is Unknown. PLoS ONE 7:45586.

@@

Carpenter, S. R., W. a Brock, J. J. Cole, J. F. Kitchell, and M. L.
Pace. 2008. Leading indicators of trophic cascades.. Ecology letters
11:128--38.

Carpenter, S. R., J. J. Cole, M. L. Pace, R. Batt, W. a Brock, T. Cline,
J. Coloso, J. R. Hodgson, J. F. Kitchell, D. a Seekell, L. Smith, and B.
Weidel. 2011. Early Warnings of Regime Shifts: A Whole-Ecosystem
Experiment. Science 1079.

@@

Carpenter, S. R., and W. a Brock. 2006. Rising variance: a leading
indicator of ecological transition.. Ecology letters 9:311--8.

@@

Chisholm, R. a, and E. Filotas. 2009. Critical slowing down as an
indicator of transitions in two-species models.. Journal of theoretical
biology 257:142--9.

@@

Hirota, M., M. Holmgren, E. H. Van Nes, and M. Scheffer. 2011. Global
resilience of tropical forest and savanna to critical transitions..
Science (New York, N.Y.) 334:232--5.

@@

Lindegren, M., V. Dakos, J. P. Gröger, A. G\textbackslash{}aa rdmark, G.
Kornilovs, S. a Otto, and C. Möllmann. 2012. Early Detection of
Ecosystem Regime Shifts: A Multiple Method Evaluation for Management
Application. PLoS ONE 7:38410.

@@

Melbourne, B. a, and A. Hastings. 2008. Extinction risk depends strongly
on factors contributing to stochasticity.. Nature 454:100--3.

@@

Renne, P. R., A. L. Deino, F. J. Hilgen, K. F. Kuiper, D. F. Mark, W. S.
Mitchell, L. E. Morgan, R. Mundil, and J. Smit. 2013. Time scales of
critical events around the Cretaceous-Paleogene boundary.. Science (New
York, N.Y.) 339:684--7.

Rietkerk, M., J. van de Koppel, S. C. Dekker, and P. C. de Ruiter. 2004.
Self-organized patchiness and catastrophic shifts in ecosystems..
Science (New York, N.Y.) 305:1926--9.

@@

Scheffer, M., S. R. Carpenter, T. M. Lenton, J. Bascompte, W. Brock, V.
Dakos, J. van de Koppel, I. A. van de Leemput, S. a Levin, E. H. van
Nes, M. Pascual, and J. Vandermeer. 2012. Anticipating Critical
Transitions. Science 338:344--348.

@@

>>>>

Many natural systems exhibit regime shifts - rapid changes in the state
and conditions of system behavior. Examples of such shifts include lake
eutrophication (Carpenter et al. 1999), algal overgrowth of coral
systems (Mumby et al. 2007), fishery collapse (Jackson et al. 2001),
desertification of grasslands (Kéfi et al. 2007), and rapid changes in
climate (Dakos et al. 2008, Lenton et al. 2009). Such dramatic shifts
have the potential to impact ecosystem health and human well-being.
Thus, it is important to develop strategies for adaptation, mitigation,
and avoidance of such shifts.

The idea that complex systems such as ecosystems could change suddenly
and without warning goes back to the 1960s (Lewontin 1969, Holling 1973,
May 1977). Such early work revealed that even simple models with the
appropriate nonlinearities were capable of unpredictable behavior. The
only way to predict the transition was to have the right model -- and
that meant having already had the chance to observe the transition. One
cogent early example (Ludwig et al. 1978) demonstrated how knowledge of
the forms and time scales of interactions among insects, birds, and
trees could lead to a qualitative model that essentially predicted the
possibility of regime shifts.

@@

Carpenter et al. (2011) provide a larger-scale example in which a lake
ecosystem is manipulated towards a sudden transition through the
introduction of a predator, while a neighboring experimental lake
provides a control. In this and similar lake systems, bifurcation is
thought to be driven in part by trophic interactions where adult fish
prey on the competitors of their juveniles (Carpenter and Kitchell 1996,
Walters and Kitchell 2001, Carpenter et al. 2008) which leads to a
saddle-node bifurcation. While the underlying dynamics of a whole lake
ecosystem are less tractable than the laboratory controlled chemoststats
of microorganisms, the system is understood well enough to anticipate
that a sudden transition can be induced under the intended manipulation.
Like the laboratory examples, this helps eliminate the options outside
the circle ``bifurcations,'' in Figure 1. The observed warning signals
then place it in the center of the diagram.

@@

Not all regime shifts are rapid. Some systems undergo bifurcations
between qualitatively different, but quantitatively similar regimes.
These transitions may be reversible. In a management setting, such
qualitative changes may be insignificant, so warning signals that detect
such transitions may be effective ``false positives.''

@@

Large, rapid changes in external conditions will result in rapid changes
in ecological system dynamics. For instance, rapid changes in North
American vegetation at the start of the Bølling-Allerød and end of the
Younger Dryas period are thought to be responses to similarly large,
rapid changes in climate (Williams et al. 2011). Doney and Sailley
(2013) interpret a recent analysis by Di Lorenzo and Ohman (2013) as
demonstrating that what were previously thought of as regime shifts in
krill dynamics in the Pacific ocean (Hare and Mantua 2000) could
actually be explained by a close coupling to the external forcing of El
Nino environmental dynamics through the Pacific Decadal Oscillation
(PDO). Schooler et al. (2011) found that lakes with the invasive plant
\emph{Salvaniai molesta} and herbivorous weevils alternated between low-
and high-\emph{Salvnia} states driven by disturbances from regular
external flooding events. These examples highlight cases that involve
critical transitions between regimes under circumstances that do not
permit the discovery of early warning signals, as CSD is not anticipated
under these mechanisms.

@@

A wide variety of statistical summary indicators have been examined as
potential detectors of CSD. The most common are variance and
autocorrelation. Others include skewness (Guttal and Jayaprakash 2008a)
and conditional heteroscedasticity (Seekell et al. 2011). These
statistics are typically calculated on sliding windows of time-series
data and tested formally or informally for trends. The relative power of
these tests varies considerably with context; no indicator has
consistently outperformed others (Dakos et al. 2011b, 2012, Lindegren et
al. 2012, Perretti and Munch 2012). Also, measuring these indicators
requires making sometimes arbitrary calculations. For instance, the
power of lag-1 autocorrelation to detect a regime shift may be modified
by changing methods of data aggregation, de-trending, changing sliding
window length, filtering signal bandwidth (Lenton et al. 2012). These
choices may be optimized when enough calibration data is available, as
Lenton et al. (2012) were able to do with several sets of paleoclimate
data. However, such calibration may not be possible with many ecological
datasets. Multiple-method (Lindegren et al. 2012) and composite indices
(Drake and Griffen 2010) have been proposed, but their power relative to
other indicators is unknown.

@@

Thus, establishing the saddle node mechanism is a necessary condition of
using CSD as a warning signal. This can be done via manipulation in
simple experimental systems (Veraart et al. 2012, Dai et al. 2012), but
this is impractical in most natural systems. Another approach is to
assume the saddle-node mechanism applies to a limited set of systems
that have well-studied examples, such as lakes undergoing eutrophication
(Scheffer et al. 2001), lakes with `trophic-triangle' cascade mechanisms
(Carpenter and Kitchell 1996, Walters and Kitchell 2001, Carpenter et
al. 2008), forest/savannah transitions (Staver et al. 2011, Hirota et
al. 2011), and rangeland transitions (Walker1993; Anderies et al. 2002).
Fitting simplified saddle-node models to past regime shifts (Boettiger
and Hastings 2012a) in less well-understood systems may provide evidence
for the mechanism. However, care must be taken to specify sufficient
alternative models.

@@

The future of early warning signals lies in the uncharted territory. For
certain classes of transitions, such as stochastically-driven regime
shifts, prediction may not be possible. In such cases, management
options include optimizing outcomes despite the possibility of regime
shifts, or possibly taking actions to reduce the long-term probability
of regime shifts, despite short-term unpredictability. Likewise, regime
shifts driven by external perturbation or strong forcing are not
predictable \emph{if} the scope of management does not include the
external causes. Proper scoping of the management problem can avoid this
situation (Fischer et al. 2009, Alliance 2010, Polasky et al. 2011).
More research is needed in methods of distinguishing such cases from
those in which early detection may be possible.

@@

Anderies, J. M., M. A. Janssen, and B. H. Walker. 2002. Grazing
Management, Resilience, and the Dynamics of a Fire-driven Rangeland
System. Ecosystems 5:23--44.

@@

Barnosky, A. D., E. A. Hadly, J. Bascompte, E. L. Berlow, J. H. Brown,
M. Fortelius, W. M. Getz, J. Harte, A. Hastings, P. A. Marquet, N. D.
Martinez, A. Mooers, P. Roopnarine, G. Vermeij, J. W. Williams, R.
Gillespie, J. Kitzes, C. Marshall, N. Matzke, D. P. Mindell, E. Revilla,
and A. B. Smith. 2012. Approaching a state shift in Earth's biosphere.
Nature 486:52--58.

@@

Brock, W. A., and S. R. Carpenter. 2012. Early Warnings of Regime Shift
When the Ecosystem Structure Is Unknown. PLoS ONE 7:45586.

@@

Carpenter, S. R., W. A. Brock, J. J. Cole, J. F. Kitchell, and M. L.
Pace. 2008. Leading indicators of trophic cascades.. Ecology letters
11:128--38.

Carpenter, S. R., J. J. Cole, M. L. Pace, R. Batt, W. A. Brock, T.
Cline, J. Coloso, J. R. Hodgson, J. F. Kitchell, D. a Seekell, L. Smith,
and B. Weidel. 2011. Early Warnings of Regime Shifts: A Whole-Ecosystem
Experiment. Science 1079.

@@

Carpenter, S. R., and W. A. Brock. 2006. Rising variance: a leading
indicator of ecological transition.. Ecology letters 9:311--8.

@@

Chisholm, R. A., and E. Filotas. 2009. Critical slowing down as an
indicator of transitions in two-species models.. Journal of theoretical
biology 257:142--9.

@@

Hirota, M., M. Holmgren, E. H. Van Nes, and M. Scheffer. 2011. Global
resilience of tropical forest and savanna to critical transitions..
Science 334:232--5.

@@

Lindegren, M., V. Dakos, J. P. Gröger, A. G\textbackslash{}aa rdmark, G.
Kornilovs, S. A. Otto, and C. Möllmann. 2012. Early Detection of
Ecosystem Regime Shifts: A Multiple Method Evaluation for Management
Application. PLoS ONE 7:38410.

@@

Melbourne, B. A., and A. Hastings. 2008. Extinction risk depends
strongly on factors contributing to stochasticity.. Nature 454:100--3.

@@

Renne, P. R., A. L. Deino, F. J. Hilgen, K. F. Kuiper, D. F. Mark, W. S.
Mitchell, L. E. Morgan, R. Mundil, and J. Smit. 2013. Time scales of
critical events around the Cretaceous-Paleogene boundary.. Science
339:684--7.

Rietkerk, M., J. van de Koppel, S. C. Dekker, and P. C. de Ruiter. 2004.
Self-organized patchiness and catastrophic shifts in ecosystems..
Science 305:1926--9.

@@

Scheffer, M., S. R. Carpenter, T. M. Lenton, J. Bascompte, W. A. Brock,
V. Dakos, J. van de Koppel, I. A. van de Leemput, S. A. Levin, E. H. van
Nes, M. Pascual, and J. Vandermeer. 2012. Anticipating Critical
Transitions. Science 338:344--348.

@@

Walters, C., and J. F. Kitchell. 2001. Cultivation/depensation effects
on juvenile survival and recruitment: implications for the theory of
fishing. Canadian Journal of Fisheries and Aquatic Sciences 58:39--50.
